**OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units**

### 2017-02-03 submission/response_to_reviewers.md

---
title: ""
output:
  pdf_document:
    includes:
      in_header: header.tex
fontsize: 11pt
geometry: margin=1.0in
---

The line numbers referred to in our response to reviewers correspond to those in the version of the manuscript with the "track changes" noted (i.e. `track_changes.pdf`).

@@

## Reviewer 1

### Major comments
**Optimizing a criterion is a good idea. It would be useful to clarify if each of the other algorithms aims to optimize some other criterion and if so which one. There is some text about this in the discussion, but this information might work better earlier in the manuscript. It would also be helpful to see how each algorithm performs on these other metrics (e.g., minimizing FPs). You can argue that MCC is a better criterion to use, but it is also good to show if the methods that aim to optimize another criterion actually do so and how MCC-based OptiClust does with regards to these metrics. In addition, some further discussion about the relative costs of FPs and FNs in typical metagenomics applications and downstream studies would provide some context for your argument to use MCC.**

We have added text to the Introduction to describe that the furthest and nearest neighbor algorithms no not allow FPs and FNs, respectively (L83-84). We have also added text to the Discussion (L244-251) pointing out that More FPs would increase the number of OTUs while more FNs would collapse OTUs together. It is difficult to know which is worse since both alpha and beta diversity metrics are tied to the number and composition of OTUs. As we point out, this is why it is important to jointly minimize the number of FPs and FNs.

@@

**For the simulations, it would be nice to show the true values of some of the output measures (e.g., number of clusters) versus the estimated values.**

Unfortunately, this is a major problem with evaluating clustering algorithms. Except under highly contrived conditions, we do not know what the true values are for output measures such as the number of clusters. This problem is outlined in L62-88. This is why we have developed the concept of the MCC over our last several papers on the subject of OTU clustering.

**A systematic evaluation of the influence of additional variables on performance (i.e., accuracy metrics, RAM, and speed) would provide better insight into expected performance on other data sets. The analyzed data sets do cover a broad range of environments with different types of communities. Even with those data, it would be nice to see performance metrics plotted versus phylogenetic diversity, percentage of sequences in the reference database (for pre-assignment analyses), and sequencing variables (e.g., read length, quality scores). In the simulations, you could additionally introduce realistic sequencing errors evaluate how performance changes.**

As the reviewer correctly points out, there is no ideal dataset. This is why we selected four datasets that represent a variety of environments that are currently popular among microbial ecologists. We also added the simulated datasets (i.e. even and staggered) since these have been used in prior studies evaluating clustering methods. Given that the nature of the algorithm is to optimize the MCC coefficient (or any other coefficient) and we have demonstrated that it is effective at doing this, we don't think it is necessary to add additional datasets to reinfoce the point. To the reviewer's suggestion, we have added the number of OTUs that were found using the OptiClust algorithm and the number of pairwise distances that were equal to and smaller than 0.03 to the summary statistics in Table 1. Unfortunately, some of the datasets are too large and the sequencing data too noisy to permit the construction of the phylogenetic trees that would be needed to calculate the phylogenetic diversity for each study. As for comparing to percentage of sequences in a reference database, we feel that this is would be a misleading point as we have previously shown that reference-based methods are inferior to *de novo* clustering methods (L47; reference added) and the identity of the sequences does not affect the performance of the algorithm (i.e. we could have used synthetic DNA sequences and communities and reached the same conclusion). Finally, the the point regarding read length, quality scores, and sequencing errors we do feel that these analyses are beyond the scope of the current paper. OptiClust only uses pairwise distances as inputs to the clustering and data quality is a very important, but different problem. Given the data provided in Table 1 and Figure 2, we have demonstrated the effect of increasing the number of distances on the performance of the algorithm.

@@

**This may be beyond the scope, but it would be very interesting to see the performance of OptiClust with other criteria as the target of optimization. In other words, how much of the performance is due to targeting MCC as the criterion versus the algorithm itself as an alternative to hierarchical clustering. You could investigate this by comparing the OptiClust algorithm minimizing FPs, for example, as compared to hierarchical clustering minimizing FPs. If this analysis is not performed, I recommend discussing that the algorithm for building clusters and the criterion that is being optimized are evaluated together in your analyses and not disentangled.**

Because both reviewers inquired about this point, we added two supplementary figures. The first demonstrates the performance of the algorithm when we optimized clustering based on a variety of other metrics (Figure S1). The second, demonstrates how the MCC values improve when optimizing for the MCC value with each successive iteration (Figure S2).

@@

**Line 138: With partial convergence, does the output include the last value for change in MCC?**

Yes, the MCC values are provided as part of the output of running the command.

**Lines 140-150: Mention if agglomerative hierarchical methods are also better than divisive (cite literature or own data).**

We have added a sentence to the Introduction (L52) indicating that all of the *de novo* methods used by microbial ecologists are agglomerative.

@@

**Line 170: Is disk or CPU also useful to measure?**

In our experience users are limited by RAM and want the methods to complete as quickly as possible. This is shown by the failure of the USEARCH methods to load the datasets in the available RAM or other algorithms to finish in 50 hrs with 48 GB of RAM. The availability of disk space is rarely a problem.

**Lines 182-185: Clarify that this is assuming the pairwise similarities are similar to the analyzed data set. Can you say something about what happens as diversity increases or decreases?**

We have added text (L175-177) to indicate that speed was proportionate to the number of distances smaller than 0.03. It does not necessarily appear to be related to the number of sequences or the number of OTUs.

@@

**Line 187: Remind the reader that this is “open reference”**

A sentence has been added to clarify the differences between our method and open reference clustering (L198-201). Briefly, in open reference clustering sequences are classified and any that do not map to a reference are clustered using a de novo approach. In our approach, all sequences are classified (not mapped) to a broad taxonomic level and then all sequences are clustered using a de novo approach.

@@

**On page 6 and 7 the authors compare the speed of OptiClust to other methods; however, I do not think Figure 1 really conveys this, mostly due to the log-scaled y-axis for the time panel.  Whilst I recognise why the log-scale has been used, the authors may consider a different way to visually present the data that better conveys the message "The OptiClust algorithm was considerably faster than the hierarchical algorithms and somewhat slower than the heuristic-based algorithms"**

We tried several methods for improving the visualization. First, we made all speed values relative to those in the first column. The problem with this was that the range of values was still considerable and obscured the ability to see the differences. Second, we added grid lines that overlapped with the values from the first column and extended to the right along the plot. This ended up making the figure look overly busy. Third, we attempted to group the methods within the sample types (we had grouped the samples within the methods type), but it was difficult to clearly represent the 13 different methods and 6 samples efficiently. Ultimately, we decided to add a thin gridline at the major y-axis values to give the reader something to anchor their perspective as they looked across the plot to compare the different methods.

**The comparison of RAM used (at the top of page 7) may also be better demonstrated visually.**

At L120-123, our goal wasn't to compare RAM usage between methods. We merely wanted to point out that the USEARCH-based approaches could not handle large datasets without purchasing the 64-bit version of the software and that OTUCLUST and SumClust could not store the data within 45 GB of RAM.

@@

**It wasn't clear to me which software implementation was used for "Nearest Neighbour" and "Furthest Neighbour".**

These were run using the implementations found in mothur (L313).

@@

**The authors describe the potential implications of optimising a metric other than MCC - I wonder if they have actually done this?**

We have added text and two supplemental figures to the manuscript in response to this comment and that of the first reviewer. We addressed this point above.

### 2017-02-09 submission/Westcott_OptiClust_mSphere_2016.Rmd
Finalize documents

***OptiClust algorithm.***
The OptiClust algorithm uses the pairs of sequences that are within a desired threshold of each other (e.g. 0.03), a list of all sequence names in the dataset, and the metric that should be used to assess clustering quality. A detailed description of the algorithm is provided for a toy dataset in the Supplementary Material. Briefly, the algorithm starts by placing each sequence either within its own OTU or into a single OTU. The algorithm proceeds by interrogating each sequence and re-calculating the metric for the cases where the sequence stays in its current OTU, is moved to each of the other OTUs, or is moved into a new OTU. The location that results in the best clustering quality indicates whether the sequence should remain in its current OTU or be moved to a different or new OTU. Each iteration consists of interrogating every sequence in the dataset. Although numerous options are available for optimizing the clusters and for assessing the quality of the clusters within the mothur-based implementation of the algorithm (e.g. sensitivity, specificity, accuracy, F1-score, etc.), the default metric for optimization and assessment is MCC because it includes all four parameters from the confusion matrix (Figure S1). The algorithm continues until the optimization metric stabilizes or until it reaches a defined stopping criteria.

@@

***OptiClust-generated OTUs are more robust than those from other methods.***
To evaluate the OptiClust algorithm and compare its performance to other algorithms, we utilized six datasets including two synthetic communities and four previously published large datasets generated from soil, marine, human, and murine samples (Table 1). When we seeded the OptiClust algorithm with each sequence in a separate OTU and ran the algorithm until complete convergence, the MCC values averaged `r round(mcc_rel_an, 1)` and `r round(mcc_rel_vdgc, 1)`% higher than the OTUs using average neighbor and distance-based greedy clustering (DGC) with VSEARCH, respectively (Figure 1). The number of OTUs formed by the various methods was negatively correlated with their MCC value ($\rho$=`r round(mcc_sobs_cor$estimate, 2)`; p`r p<-mcc_sobs_cor$p.value; ifelse(p<0.001, "<0.001", paste0("=", p))`). The OptiClust algorithm was considerably faster than the hierarchical algorithms and somewhat slower than the heuristic-based algorithms. Across the six datasets, the OptiClust algorithm was `r round(secs_rel_an, 1)`-times faster than average neighbor and just as fast as DGC with VSEARCH. The human dataset was a challenge for a number of the algorithms. OTUCLUST and SumaClust were unable to cluster the human dataset in less than 50 hours and the average neighbor algorithm required more than 45 GB of RAM. The USEARCH-based methods were unable to cluster the human data using the 32-bit free version of the software that limits the amount of RAM to approximately 3.5 GB. These data demonstrate that OptiClust generated significantly more robust OTU assignments than existing methods across a diverse collection of datasets with performance that was comparable to popular methods.

@@

***Implementation of clustering algorithms.***
In addition to the OptiClust algorithm we evaluated ten different *de novo* clustering algorithms. These included three hierarchical algorithms, average neighbor, nearest neighbor, and furthest neighbor, which are implemented in mothur (v.1.39.0) [@Schloss2009b]. Seven heuristic methods were also used including abundance-based greedy clustering (AGC) and (distance-based greedy clustering) DGC as implemented in USEARCH (v.6.1) [@Edgar2010] and VSEARCH (v.2.3.3) (@Rognes2016], OTUCLUST (v.0.1) [@Albanese2015], SumaClust (v.1.0.20), and Swarm (v.2.1.9) [@Mah2014]. With the exception of Swarm each of these methods uses distance-based thresholds to report OTU assignments. We also evalauted the ability of OptiClust to optimize to metrics other than MCC. These included accuracy, F1-score, negative predictive value, positive predictive value, false discovery rate, senitivity, specificity, the sum of TPs and TNs, the sum of FPs and FNs, and the number of FNs, FPs, TNs, and TPs (Figure S1).

@@

>>>>

***OptiClust algorithm.***
The OptiClust algorithm uses the pairs of sequences that are within a desired threshold of each other (e.g. 0.03), a list of all sequence names in the dataset, and the metric that should be used to assess clustering quality. A detailed description of the algorithm is provided for a toy dataset in the Supplementary Material. Briefly, the algorithm starts by placing each sequence either within its own OTU or into a single OTU. The algorithm proceeds by interrogating each sequence and re-calculating the metric for the cases where the sequence stays in its current OTU, is moved to each of the other OTUs, or is moved into a new OTU. The location that results in the best clustering quality indicates whether the sequence should remain in its current OTU or be moved to a different or new OTU. Each iteration consists of interrogating every sequence in the dataset. Although numerous options are available for optimizing the clusters and for assessing the quality of the clusters within the mothur-based implementation of the algorithm (e.g. sensitivity, specificity, accuracy, F1-score, etc.), the default metric for optimization and assessment is MCC because it includes all four parameters from the confusion matrix (Figure S1; Table S1). The algorithm continues until the optimization metric stabilizes or until it reaches a defined stopping criteria.

@@

***OptiClust-generated OTUs are more robust than those from other methods.***
To evaluate the OptiClust algorithm and compare its performance to other algorithms, we utilized six datasets including two synthetic communities and four previously published large datasets generated from soil, marine, human, and murine samples (Table 1). When we seeded the OptiClust algorithm with each sequence in a separate OTU and ran the algorithm until complete convergence, the MCC values averaged `r round(mcc_rel_an, 1)` and `r round(mcc_rel_vdgc, 1)`% higher than the OTUs using average neighbor and distance-based greedy clustering (DGC) with VSEARCH, respectively (Figure 1; Table S1). The number of OTUs formed by the various methods was negatively correlated with their MCC value ($\rho$=`r round(mcc_sobs_cor$estimate, 2)`; p`r p<-mcc_sobs_cor$p.value; ifelse(p<0.001, "<0.001", paste0("=", p))`). The OptiClust algorithm was considerably faster than the hierarchical algorithms and somewhat slower than the heuristic-based algorithms. Across the six datasets, the OptiClust algorithm was `r round(secs_rel_an, 1)`-times faster than average neighbor and just as fast as DGC with VSEARCH. The human dataset was a challenge for a number of the algorithms. OTUCLUST and SumaClust were unable to cluster the human dataset in less than 50 hours and the average neighbor algorithm required more than 45 GB of RAM. The USEARCH-based methods were unable to cluster the human data using the 32-bit free version of the software that limits the amount of RAM to approximately 3.5 GB. These data demonstrate that OptiClust generated significantly more robust OTU assignments than existing methods across a diverse collection of datasets with performance that was comparable to popular methods.

@@

***Implementation of clustering algorithms.***
In addition to the OptiClust algorithm we evaluated ten different *de novo* clustering algorithms. These included three hierarchical algorithms, average neighbor, nearest neighbor, and furthest neighbor, which are implemented in mothur (v.1.39.0) [@Schloss2009b]. Seven heuristic methods were also used including abundance-based greedy clustering (AGC) and (distance-based greedy clustering) DGC as implemented in USEARCH (v.6.1) [@Edgar2010] and VSEARCH (v.2.3.3) (@Rognes2016], OTUCLUST (v.0.1) [@Albanese2015], SumaClust (v.1.0.20), and Swarm (v.2.1.9) [@Mah2014]. With the exception of Swarm each of these methods uses distance-based thresholds to report OTU assignments. We also evalauted the ability of OptiClust to optimize to metrics other than MCC. These included accuracy, F1-score, negative predictive value, positive predictive value, false discovery rate, senitivity, specificity, the sum of TPs and TNs, the sum of FPs and FNs, and the number of FNs, FPs, TNs, and TPs (Figure S1; Table S1).

@@

**Table S1. Summary of the average number of true positives, true negatives, false positives, false negatives and the resulting Matthews correlation coefficient for each of the clustering methods that were analyzed in this study for each of the six datasets.** Blank values indicate that those conditions could not be completed in 50 hours with 45 GB of RAM.
